How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants

### 2022-10-09 ms/ms.tex
minor changes to text

\title{Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants}

@@

\setstretch{1.7}
\maketitle

\bigskip

\textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, figure 6, table 1, online appendix A (including Notes A1, notes A2, notes A3, figure A1, figure A2, figure A3, table A1, table A2, and table A3).

\bigskip

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, packing limits, phylogenetic comparative methods, stomata

\bigskip

\textit{Manuscript type}: Major Article

\bigskip

\textit{Word Count}: 4,704

\bigskip

@@

\newpage{}

@@

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the importance of functional and developmental constraints on phenotypyic evolution. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially different selective constraints because of leaf asymmetry in light capture, gas exchange, and other features. Independent evolution of stomatal traits on each surface imply that functional and developmental constraints alone likely do not explain trait covariance. Packing limits on how many stomata can fit into a finite epidermis and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. The simple geometry of the planar leaf surface and knowledge of stomatal development make it possible to derive equations for phenotypic (co)variance caused by these constraints and compare them with data. We analyzed evolutionary covariance between stomatal density and length in amphistomatous leaves from 236 phylogenetically independent contrasts using a robust Bayesian model. Stomatal anatomy on each surface diverges partially independently, meaning that packing limits and developmental integration are not sufficient to explain phenotypic (co)variation. Hence, selective constraints, which require an adaptive explanation, likely contribute to (co)variation in ecologically important traits like stomata. We show how it is possible to evaluate the contribution of different constraints by deriving expected patterns of (co)variance and testing them using similar but separate tissues, organs, or sexes.\\

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for populations to ascend a fitness peak in multivariate trait space (\protect\hyperlink{ref-lewontin_adaptation_1978}{Lewontin 1978}). If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years (\protect\hyperlink{ref-schluter_adaptive_1996}{Schluter 1996}). Covariance is ``a rough local measure of the strength of constraint'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}) that can be broken down into functional, developmental, and selective constraints. Functional constraints are ``limitations imposed by time, energy, or the laws of physics'' (\protect\hyperlink{ref-arnold_constraints_1992}{Arnold 1992}). In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits (\protect\hyperlink{ref-raup_geometric_1966}{Raup 1966}; \protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}). Within the space of possible phenotypes, developmental constraints can ``bias\ldots the production of variant phenotypes or {[}place{]} a limitation on phenotypic variability'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}). For example, Fibonacci phyllotaxis may arise from a packing constraint of primordia on the developing apex (\protect\hyperlink{ref-mitchison_phyllotaxis_1977}{Mitchison 1977}; \protect\hyperlink{ref-reinhardt_law_2022}{Reinhardt and Gola 2022}; but see \protect\hyperlink{ref-niklas_role_1988}{Niklas 1988} for an adaptive explanation). Natural selection constrains phenotypic variation by preventing maladaptive forms from evolving. Selection causes trait covariation because `missing' trait combinations are maladaptive. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints (\protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}, \protect\hyperlink{ref-mcghee_geometry_2007}{2007}; \protect\hyperlink{ref-olson_plant_2019}{Olson 2019}). If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test for selective constraint. Selective constraint does not mean that populations do not respond to selection, but rather that limits on phenotypic variation arise because selection never favors certain trait values.

@@

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or too poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important.

\hypertarget{the-adaptive-significance-and-ecological-distribution-of-variation-in-stomatal-anatomy}{%
\subsection{The adaptive significance and ecological distribution of variation in stomatal anatomy}\label{the-adaptive-significance-and-ecological-distribution-of-variation-in-stomatal-anatomy}}

@@

Two major features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as ``inverse size-density scaling'' and ``bimodal stomatal ratio'' (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata (\protect\hyperlink{ref-weiss_untersuchungen_1865}{Weiss 1865}; \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}). Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO\(_2\) and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO\(_2\) for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO\(_2\). Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO\(_2\) concentration gradient around one stomate merges with that of its neighbor (\protect\hyperlink{ref-zeiger_stomatal_1987}{Zeiger, Farquhar, and Cowan 1987}; \protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015}), incur metabolic costs (\protect\hyperlink{ref-deans_optimization_2020}{Deans et al. 2020}), and/or risk hydraulic failure (\protect\hyperlink{ref-henry_stomatal_2019}{Henry et al. 2019}). The distribution of stomatal size and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}) further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

@@

Given the significance of stomata for plant function and global vegetation modeling (\protect\hyperlink{ref-berry_stomata:_2010}{Berry, Beerling, and Franks 2010}), we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecifc variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}) for the species' mean, not an individual leaf.

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}), the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}). For example, larger genomes increase meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), which sets a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration on stomatal anatomy (\protect\hyperlink{ref-beaulieu_genome_2008}{Beaulieu et al. 2008}; \protect\hyperlink{ref-simonin_genome_2018}{Simonin and Roddy 2018}; \protect\hyperlink{ref-roddy_scaling_2020}{Roddy et al. 2020}). Developmental integration in this case would not necessarily hinder adaptive evolution if the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}), then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

Muir (\protect\hyperlink{ref-muir_making_2015}{2015}) derived general conditions in which bimodality arises because adaptive optima are restricted to separate regimes, but this model has not been tested. An alternative hypothesis is that stomatal traits on the ab- and adaxial surfaces are developmentally integrated because stomatal development is regulated the same way on each surface. In hypostomatous leaves, stomatal development is turned off in the adaxial surface. In amphistomatous leaves, stomatal development proceeds on both surfaces, but evolutionary changes in stomatal development affect traits on both surfaces because they are tethered by a shared developmental program. This is a developmental constraint because the fact that stomatal development is the same on each surface constrains the type of variation available for selection. Developmental integration would lead to a bimodal trait distribution because leaves would either be hypostomatous (stomatal ratio equal to 0) or have similar densities on each surface (stomatal ratio approximately 0.5). To our knowledge, this hypothesis has not been put forward in the literature.

@@

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and/or developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selective constraints likely contribute to covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection (\protect\hyperlink{ref-barrett_sexual_2013}{Barrett and Hough 2013}). We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

The second hypothesis is the cell size mediates developmental integration between stomatal size and density. If developmental integration is the primary reason for inverse size-density scaling, then amphistomatous leaves will exhibit identical size-density scaling on each surface. If the stomatal size and density scale differently on each surface, this implies that they can evolve independently and that selective constraints likely explain some of their covariance. Furthermore, we predicted that divergence in genome size, which is strongly associated with meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), would covary with stomatal size and density similarly on each surface.

@@

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy (\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}). We supplemented these by searching Clarivate \emph{Web of Science} for ``guard cell length'' because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through opportunistic discovery. The final data set contained 5104 observations of stomatal density and length from 1242 taxa and 38 primary studies (Table \ref{tab:sources}). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included some unpublished data. Stomatal size data were collected on grass species described in Pathare, Koteyeva, and Cousins (\protect\hyperlink{ref-pathare_increased_2020}{2020}). We also included unpublished data on 14 amphistomatous wild tomato species (\emph{Solanum} sect. \emph{Lycopersicum} and sect. \emph{Lycopersicoides}) grown in pots under outdoor summer Mediterranean conditions as described in Muir, Galmés, and Conesa (\protect\hyperlink{ref-muir_unpublished_2022}{2022}). We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three 0.571 mm\(^2\) fields of view and divided by the total area to estimate density. We measured the average guard cell length of 60 stomata, 20 per field of view, to estimate stomatal size. The data set is publicly available as an \emph{R} package \textbf{ropenstomata} (\url{https://github.com/cdmuir/ropenstomata}). We collected data on genome size from the Angiosperm DNA C-values database (\protect\hyperlink{ref-leitch_angiosperm_2019}{Leitch et al. 2019}; \protect\hyperlink{ref-pellicer_plant_2020}{Pellicer and Leitch 2020}). When multiple ploidy levels were available for a taxon, we chose the lowest one for consistency. All data will be deposited on Dryad and archived on Zenodo upon publication.

@@

We tested the packing limit hypothesis by estimating the effect of \(f_S\) on \(\textrm{Var}[\Delta~\textrm{log(trait)}]\) for each trait and leaf surface. If there is an upper bound on \(f_S\), we predict the effect of \(f_S\) on \(\textrm{Var}[\Delta~\textrm{log(trait)}]\) will be \(<0\). Specifically, the 95\% HPD intervals should not include 0. Further, the coefficient should be the same on each surface, so the 95\% HPD intervals for difference should encompass 0.

@@

\hypertarget{divergence-in-stomatal-traits-slows-as-f_s-increases}{%
\subsection{\texorpdfstring{Divergence in stomatal traits slows as \(f_S\) increases}{Divergence in stomatal traits slows as f\_S increases}}\label{divergence-in-stomatal-traits-slows-as-f_s-increases}}

The variance in trait divergence decreases as the fraction of epidermal area allocated to stomata, \(f_S\), increases (Figs. \ref{fig:h3}, \ref{fig:fs-sigma}). The effect of \(f_S\) was strongest for \(D_\text{ad}\) and 95\% HPD intervals did not overlap 0 for 3 of 4 comparisons (Table \ref{tab:h3output}). Variance in divergence for \(D_\text{ad}\) declined more rapidly with \(f_S\) than that for \(D_\text{ab}\) (difference and 95\% HPD interval in slope, log-link scale: -10.7 {[}-17.2,-4.2{]}). Variance in length divergence declined similarly with \(f_S\) on both surfaces (difference and 95\% HPD interval in slope, log-link scale: -2.8 {[}-9.6,4{]}).

@@

\hypertarget{adaxial-stomatal-density-is-more-variable-but-size-density-covariance-is-similar-on-both-surfaces}{%
\subsection{Adaxial stomatal density is more variable, but size-density covariance is similar on both surfaces}\label{adaxial-stomatal-density-is-more-variable-but-size-density-covariance-is-similar-on-both-surfaces}}

@@

\hypertarget{genome-size-is-associated-with-stomatal-length-on-both-surfaces}{%
\subsection{Genome size is associated with stomatal length on both surfaces}\label{genome-size-is-associated-with-stomatal-length-on-both-surfaces}}

@@

>>>>

\title{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants}

@@

\setstretch{1.7}
\begin{flushleft}
{\Large
\textbf\newline{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants}
}
\newline
\\
Christopher D. Muir\textsuperscript{1,*},
Miquel \`{A}ngel Conesa\textsuperscript{2},
Jeroni Galm\'{e}s\textsuperscript{2},
Varsha S. Pathare\textsuperscript{3},
Patricia Rivera\textsuperscript{4},
Rosana López Rodríguez\textsuperscript{5},
Teresa Terrazas\textsuperscript{4},
Dongliang Xiong\textsuperscript{6}
\\
\bigskip
\bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\
\bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\
\bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\
\bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\
\bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\
\bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\
\bigskip
* cdmuir@hawaii.edu

\end{flushleft}

@@

@@

Quantifying the relative contribution of functional and developmental constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. Alternatively, selection can limit phenotypic (co)variation if some trait combinations are generally maladaptive. The anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the importance of functional and developmental constraints on phenotypyic evolution. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially different selective pressures because of leaf asymmetry in light capture, gas exchange, and other features. Independent evolution of stomatal traits on each surface imply that functional and developmental constraints alone likely do not explain trait covariance. Packing limits on how many stomata can fit into a finite epidermis and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. The simple geometry of the planar leaf surface and knowledge of stomatal development make it possible to derive equations for phenotypic (co)variance caused by these constraints and compare them with data. We analyzed evolutionary covariance between stomatal density and length in amphistomatous leaves from 236 phylogenetically independent contrasts using a robust Bayesian model. Stomatal anatomy on each surface diverges partially independently, meaning that packing limits and developmental integration are not sufficient to explain phenotypic (co)variation. Hence, (co)variation in ecologically important traits like stomata arises in part because there is a limited range of evolutionary optima. We show how it is possible to evaluate the contribution of different constraints by deriving expected patterns of (co)variance and testing them using similar but separate tissues, organs, or sexes.\\

@@

Selection should move populations toward their multivariate phenotypic optimum if traits can evolve independently and there is sufficient genetic variation. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years (\protect\hyperlink{ref-schluter_adaptive_1996}{Schluter 1996}). Covariance is ``a rough local measure of the strength of constraint'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}) that can be broken down into functional and developmental constraints. Functional constraints are ``limitations imposed by time, energy, or the laws of physics'' (\protect\hyperlink{ref-arnold_constraints_1992}{Arnold 1992}). In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits (\protect\hyperlink{ref-raup_geometric_1966}{Raup 1966}; \protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}). Within the space of possible phenotypes, developmental constraints can ``bias\ldots the production of variant phenotypes or {[}place{]} a limitation on phenotypic variability'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}). For example, Fibonacci leaf arrangement (phyllotaxis) may arise from a packing constraint of primordia on the developing apex (\protect\hyperlink{ref-mitchison_phyllotaxis_1977}{Mitchison 1977}; \protect\hyperlink{ref-reinhardt_law_2022}{Reinhardt and Gola 2022}; but see \protect\hyperlink{ref-niklas_role_1988}{Niklas 1988} for an adaptive explanation). The adaptationist view is that all adaptive areas of trait space are occupied and that natural selection constrains phenotypic variation because maladaptive forms cannot evolve. Under this view, interspecific trait covariation evolves because `missing', maladaptive trait combinations are nonrandomly distributed in trait space. For example, a negative correlation between apparency and toxicity can arise because prey can avoid predation by being cryptic or toxic, but no prey population can exist for long if it is both apparent and palatable. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints (\protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}, \protect\hyperlink{ref-mcghee_geometry_2007}{2007}; \protect\hyperlink{ref-olson_plant_2019}{Olson 2019}). If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test adaptive explanations. This is sometimes referred to as selective constraint because selection eliminates phenotypic variation that is generally maladaptive.

@@

Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity to test for functional and developmental constraints because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selection may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selection against generally maladaptive phenotypes explains at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important.

\hypertarget{how-stomatal-anatomy-varies-and-why-it-matters}{%
\subsection{How stomatal anatomy varies and why it matters}\label{how-stomatal-anatomy-varies-and-why-it-matters}}

@@

Two major features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We refer to these features as ``inverse size-density scaling'' and ``bimodal stomatal ratio'' (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata (\protect\hyperlink{ref-weiss_untersuchungen_1865}{Weiss 1865}; \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}). Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO\(_2\) and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO\(_2\) for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO\(_2\). Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO\(_2\) concentration gradient around one stomate merges with that of its neighbor (\protect\hyperlink{ref-zeiger_stomatal_1987}{Zeiger, Farquhar, and Cowan 1987}; \protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015}), incur metabolic costs (\protect\hyperlink{ref-deans_optimization_2020}{Deans et al. 2020}), and/or risk hydraulic failure (\protect\hyperlink{ref-henry_stomatal_2019}{Henry et al. 2019}). The distribution of stomatal size and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}) further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

@@

Given the significance of stomata for plant function and global vegetation modeling (\protect\hyperlink{ref-berry_stomata:_2010}{Berry, Beerling, and Franks 2010}), we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecific variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}) for the species' mean, not an individual leaf.

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}), the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}). For example, larger genomes increase meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), which sets a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration on stomatal anatomy (\protect\hyperlink{ref-beaulieu_genome_2008}{Beaulieu et al. 2008}; \protect\hyperlink{ref-simonin_genome_2018}{Simonin and Roddy 2018}; \protect\hyperlink{ref-roddy_scaling_2020}{Roddy et al. 2020}). Developmental integration in this case would not necessarily slow adaptive evolution if the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by Franks and Beerling (\protect\hyperlink{ref-franks_maximum_2009}{2009}), then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

Muir (\protect\hyperlink{ref-muir_making_2015}{2015}) showed that bimodality can arise adaptively if fitness optima are restricted to separate regimes. Specifically, if either the fitness costs or benefits of allocating additional stomata to the upper surface are nonlinear, with the correct curvature, then there are distinct ``hypostomatous'' and ``amphistomatous'' regimes. An alternative hypothesis is that stomatal traits on the ab- and adaxial surfaces are developmentally integrated because stomatal development is regulated the same way on each surface. In hypostomatous leaves, stomatal development is turned off in the adaxial surface. In amphistomatous leaves, stomatal development proceeds on both surfaces, but evolutionary changes in stomatal development affect traits on both surfaces because they are tethered by a shared developmental program. This is a developmental constraint because the fact that stomatal development is the same on each surface constrains the type of variation available for selection. Developmental integration would lead to a bimodal trait distribution because leaves would either be hypostomatous (stomatal ratio equal to 0) or have similar densities on each surface (stomatal ratio approximately 0.5). To our knowledge, this hypothesis has not been put forward in the literature.

@@

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and/or developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selection against generally maladaptive trait combinations shapes covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection (\protect\hyperlink{ref-barrett_sexual_2013}{Barrett and Hough 2013}). We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

The second hypothesis is the cell size mediates developmental integration between stomatal size and density. If developmental integration is the primary reason for inverse size-density scaling, then amphistomatous leaves will exhibit identical size-density scaling on each surface. If the stomatal size and density scale differently on each surface, this implies that they can evolve independently and that selection against generally maladaptive trait combinations shapes their covariance. Furthermore, we predicted that divergence in genome size, which is strongly associated with meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), would covary with stomatal size and density similarly on each surface.

@@

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy (\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}). We supplemented these by searching Clarivate \emph{Web of Science} for ``guard cell length'' because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through opportunistic discovery. The final data set contained 5104 observations of stomatal density and length from 1242 taxa and 38 primary studies (Table \ref{tab:sources}). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included some unpublished data. Stomatal size data were collected on grass species described in Pathare, Koteyeva, and Cousins (\protect\hyperlink{ref-pathare_increased_2020}{2020}). We also included unpublished data on 14 amphistomatous wild tomato species (\emph{Solanum} sect. \emph{Lycopersicum} and sect. \emph{Lycopersicoides}) grown in pots under outdoor summer Mediterranean conditions as described in Muir, Galmés, and Conesa (\protect\hyperlink{ref-muir_unpublished_2022}{2022}). We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three 0.571 mm\(^2\) fields of view and divided by the total area to estimate density. We measured the average guard cell length of 60 stomata, 20 per field of view, to estimate stomatal size. The data set is publicly available as an \emph{R} package \textbf{ropenstomata} (\url{https://github.com/cdmuir/ropenstomata}). We collected data on genome size from the Angiosperm DNA C-values database (\protect\hyperlink{ref-leitch_angiosperm_2019}{Leitch et al. 2019}; \protect\hyperlink{ref-pellicer_plant_2020}{Pellicer and Leitch 2020}). When multiple ploidy levels were available for a taxon, we chose the lowest one for consistency.

@@

We tested the packing limit hypothesis by estimating whether the fraction of epidermal area allocated to stomata (\(f_S\)) effects \(\textrm{Var}[\Delta~\textrm{log(trait)}]\) for each trait and leaf surface. If there is an upper bound on \(f_S\), we predict the effect of \(f_S\) on \(\textrm{Var}[\Delta~\textrm{log(trait)}]\) will be \(<0\). Specifically, the 95\% HPD intervals should not include 0. Further, the coefficient should be the same on each surface, so the 95\% HPD intervals for difference should encompass 0.

@@

\hypertarget{empirical-results}{%
\subsection{Empirical results}\label{empirical-results}}

The variance in trait divergence decreases as the fraction of epidermal area allocated to stomata, \(f_S\), increases (Figs. \ref{fig:h3}, \ref{fig:fs-sigma}). The effect of \(f_S\) was strongest for adaxial stomatal density (\(D_\text{ad}\)) and 95\% HPD intervals did not overlap 0 for 3 of 4 comparisons (Table \ref{tab:h3output}). Variance in divergence for \(D_\text{ad}\) declined more rapidly with \(f_S\) than that for abaxial stomatal density (\(D_\text{ab}\); difference and 95\% HPD interval in slope, log-link scale: -10.7 {[}-17.2,-4.2{]}). Variance in length divergence declined similarly with \(f_S\) on both surfaces (difference and 95\% HPD interval in slope, log-link scale: -2.8 {[}-9.6,4{]}).

@@

@@

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}
